---
title: "Project_4_2064_debajit2 : Movie Recommendation"
author: Debajit Mitra
date: '2022-12-01'
output: html_document
---

***



```{r}
start_time <- Sys.time()

library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
library(knitr)
```


### Project Overview
In this project, I developed content based recommendation system(System I) and collaborative recommendation system(System II) .

* **System I:** Recommendation based on genres. Below two recommendation approaches are implemented using user's favorite genre : 
   + __Method I:__  Highly-rate movies - select the 4 or more review
   + __Method II:__ Most trendy movies - the newest movie

* **System II:** Collaborative recommendation system. Two collaborative recommendation algorithms are implemented : 
   + __User-based (UBCF):__ UBCF recommends items by finding similar users to the user with assumption that similar users will have similar taste.
   + __Item-based (IBCF):__ IBCF recommends items by finding similarities between different items in the data, using Euclidean Distance similarity measures.
   


```{r}
set.seed(2064)
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

moviesList <- read(paste0(basedir,"aggr.dat"), "::")
movies_clean <- read(paste0(basedir,"movies_clean.dat"), "::")

colnames(moviesList) = c( 'MovieID', 'AveRating', 'title', 'genres')
colnames(movies_clean) = c("MovieID", "title", "year", 'genres')
genre_list <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")
```


### Dataset
The dataset contains about 1 million ratings and the details as following:

* Number of movies: $`r nrow(movies)`$
* Number of ratings: $`r nrow(ratings)`$
* Number of users: $`r nrow(users)`$

The dataset README shows:
Genres are pipe-separated and are selected from the following genres:

* Action
* Adventure
* Animation
* Children's
* Comedy
* Crime
* Documentary
* Drama
* Fantasy
* Film-Noir
* Horror
* Musical
* Mystery
* Romance
* Sci-Fi
* Thriller
* War
* Western
	



```{r}

genres <- as.data.frame(movies[,3], stringsAsFactors=FALSE)
genres2 <- as.data.frame(tstrsplit(genres[,1], '[|]', type.convert=TRUE), stringsAsFactors=FALSE)
colnames(genres2)<- c(1:5)  

genre_list <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")


genre_matrix <- matrix(0,nrow(movies) + 1,18) 
genre_matrix[1,] <- genre_list 
colnames(genre_matrix) <- genre_list 

for (i in 1:nrow(genres2)) {
  for (c in 1:ncol(genres2)) {
    genmat_col = which(genre_matrix[1,] == genres2[i,c])
    genre_matrix[i+1,genmat_col] <- 1
  }
}

genre_matrix2 <- as.data.frame(genre_matrix[-1,], stringsAsFactors=FALSE) 
for (c in 1:ncol(genre_matrix2)) {
  genre_matrix2[,c] <- as.integer(genre_matrix2[,c])
} 

years <- as.data.frame(movies$title, stringsAsFactors=FALSE)

substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

years <- as.data.frame(substr(substrRight(substrRight(years$`movies$title`, 6),5),1,4))
colMin <- function(data) sapply(data, min, na.rm = TRUE)
colMax <- function(data) sapply(data, max, na.rm = TRUE)

min_year = colMin(years)
max_year = colMax(years)

```

The movies' year are between $`r min_year`$ to $`r max_year`$.


```{r}
binaryratings <- ratings



for (i in 1:nrow(binaryratings)){
  if (binaryratings[i,3] > 3){
    binaryratings[i,3] <- 1
  }
  else{
    binaryratings[i,3] <- -1
  }
}

binaryratings2 <- dcast(binaryratings, MovieID~UserID, value.var = "Rating", na.rm=FALSE)



for (i in 1:ncol(binaryratings2)){
  binaryratings2[which(is.na(binaryratings2[,i]) == TRUE),i] <- 0
}
binaryratings2 = binaryratings2[,-1] 

MovieIDs <- length(unique(movies$MovieID)) 
ratingMovieIDs <- length(unique(ratings$MovieID)) 
movies2 <- movies[-which((movies$MovieID %in% ratings$MovieID) == FALSE),]
rownames(movies2) <- NULL

genre_matrix3 <- genre_matrix2[-which((movies$MovieID %in% ratings$MovieID) == FALSE),]
rownames(genre_matrix3) <- NULL

result = matrix(0,18,nrow(users)) 
for (c in 1:ncol(binaryratings2)){
  for (i in 1:ncol(genre_matrix3)){
    result[i,c] <- sum((genre_matrix3[,i]) * (as.numeric(binaryratings2[,c]))) 
  }
}

for (c in 1:ncol(result)){
  for (i in 1:nrow(result)){
    if (result[i,c] < 0){
      result[i,c] <- 0
    }
    else {
      result[i,c] <- 1
    }
  }
}

ratingmat <- dcast(ratings, UserID~MovieID, value.var = "Rating", na.rm=FALSE)
ratingmat <- as.matrix(ratingmat[,-1]) 

ratingmat <- as(ratingmat, "realRatingMatrix")

similarity_users <- similarity(ratingmat[1:4, ], 
                               method = "cosine", 
                               which = "users")
as.matrix(similarity_users)

```



### Exploring Similarity Data

```{r echo=TRUE}

image(as.matrix(similarity_users), main = "User similarity")

```


```{r}


similarity_items <- similarity(ratingmat[, 1:4], method =
                                 "cosine", which = "items")
as.matrix(similarity_items)

```


```{r echo=TRUE}

image(as.matrix(similarity_items), main = "Item similarity")

```


### Most Viewed Movies Visualization 

```{r}

vector_ratings <- as.vector(ratingmat@data)
unique(vector_ratings) 

table_ratings <- table(vector_ratings) 
table_ratings


```


```{r echo=TRUE}
vector_ratings <- vector_ratings[vector_ratings != 0] 
vector_ratings <- factor(vector_ratings)

qplot(vector_ratings) + 
  ggtitle("Distribution of the ratings")

```


```{r}


views_per_movie <- colCounts(ratingmat) 
table_views <- data.frame(movie = names(views_per_movie), views = views_per_movie) 
table_views <- table_views[order(table_views$views, decreasing = TRUE), ] 



```


```{r echo=TRUE}


#Visualizing the matrix:
image(ratingmat, main = "Heatmap of the rating matrix") # hard to read-too many dimensions
image(ratingmat[1:10, 1:15], main = "Heatmap of the first rows and columns")
image(ratingmat[rowCounts(ratingmat) > quantile(rowCounts(ratingmat), 0.99),
                 colCounts(ratingmat) > quantile(colCounts(ratingmat), 0.99)], 
      main = "Heatmap of the top users and movies")


#Normalize the data
ratingmat_norm <- normalize(ratingmat)
image(ratingmat_norm[rowCounts(ratingmat_norm) > quantile(rowCounts(ratingmat_norm), 0.99),
                colCounts(ratingmat_norm) > quantile(colCounts(ratingmat_norm), 0.99)], 
      main = "Heatmap of the top users and movies")



```

## System I

system I is content based recommendation system, where movies will be recommended to the user based on user's favorite genre.

### Method I: Highly-rate movies

  The average point of each movie is calculated and N movies are randomly picked whose average point is great or equal to 4-point in the genres. 


```{r}
  # Method 1:
  numberofmovierecommend = 5

  input_genre1 = genre_list[1]
  input_genre2 = genre_list[2]
  input_genre3 = genre_list[3]

  systemresult = subset(moviesList,AveRating>=4 & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  if (nrow(systemresult) < numberofmovierecommend){
     systemresult = subset(moviesList, grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE))
  }

  systemresult = systemresult[sample(nrow(systemresult), ifelse(nrow(systemresult)>=numberofmovierecommend,numberofmovierecommend,nrow(systemresult))),]
  systemresult


```
  

  
### Method II: Most trendy movies

  The newest movies are from year 2000 in the dataset, N movies will be randomly picked from year 2000 in that genres, 
  

```{r}
  # Method 2:
  numberofmovierecommend = 5
  trendyYear = 2000

  input_genre1 = genre_list[1]
  input_genre2 = genre_list[2]
  input_genre3 = genre_list[3]


  systemresult = subset(movies_clean, year >= trendyYear & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  if (nrow(systemresult) < numberofmovierecommend){
      systemresult = subset(movies_clean, year >= trendyYear - 1 & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  }
  systemresult = systemresult[sample(nrow(systemresult), ifelse(nrow(systemresult)>=numberofmovierecommend,numberofmovierecommend,nrow(systemresult))),]
  systemresult
```



## System II

### Create realRatingMatrix

```{r}
ratings <- as(ratings, 'realRatingMatrix')
```


### Training and Testing Data
Splitting the data into 80% training and 20% testing. 

```{r}
train_proportion <- 0.8
folds <-10
to_keep <- 15  ## given 15 items
threshold <- 0 ## ratings above 0 as the cutoff point

# split the data into the training and the test set:
e <- evaluationScheme(ratings, method="split", train=train_proportion, k=folds, given=to_keep, goodRating=threshold)
e

```
  

### Train User-based and Item-based Collaborative Filtering models

CF models are trained by below parameters:

- normalize: NULL, center or Z-score;

- method: Euclidean



#### User-Based Collaborative Filtering: Euclidean Distance

```{r}
# train UBCF Euclidean Distance models

# non-normalized
UBCF_N_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="Euclidean"))
saveRDS(UBCF_N_E, file = "model/UBCF_N_E_model.rds")
reloadModel = readRDS("model/UBCF_N_E_model.rds")

# centered
UBCF_C_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="Euclidean"))
saveRDS(UBCF_C_E, file = "model/UBCF_C_E_model.rds")
reloadModel = readRDS("model/UBCF_C_E_model.rds")

# Z-score normalization
UBCF_Z_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="Euclidean"))
saveRDS(UBCF_Z_E, file = "model/UBCF_Z_E_model.rds")
reloadModel = readRDS("model/UBCF_Z_E_model.rds")

```



```{r}
# compute predicted ratings
p1 <- predict(UBCF_N_E, getData(e, "known"), type="ratings")
p2 <- predict(UBCF_C_E, getData(e, "known"), type="ratings")
p3 <- predict(UBCF_Z_E, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_UEUC <- rbind(
  UBCF_N_E = calcPredictionAccuracy(p1, getData(e, "unknown")),
  UBCF_C_E = calcPredictionAccuracy(p2, getData(e, "unknown")),
  UBCF_Z_E = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
```



#### Item-Based Collaborative Filtering: Euclidean Distance

```{r}
#train IBCF Euclidean Distance models

# non-normalized
IBCF_N_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="Euclidean"))
saveRDS(IBCF_N_E, file = "model/IBCF_N_E_model.rds")
reloadModel = readRDS("model/IBCF_N_E_model.rds")

# centered
IBCF_C_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="Euclidean"))
saveRDS(IBCF_C_E, file = "model/IBCF_C_E_model.rds")
reloadModel = readRDS("model/IBCF_C_E_model.rds")

# Z-score normalization
IBCF_Z_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="Euclidean"))
saveRDS(IBCF_Z_E, file = "model/IBCF_Z_E_model.rds")
reloadModel = readRDS("model/IBCF_Z_E_model.rds")

```

```{r}
# compute predicted ratings
p1 <- predict(IBCF_N_E, getData(e, "known"), type="ratings")
p2 <- predict(IBCF_C_E, getData(e, "known"), type="ratings")
p3 <- predict(IBCF_Z_E, getData(e, "known"), type="ratings")


# aggregate the performance statistics
error_IEUC <- rbind(
  IBCF_N_E = calcPredictionAccuracy(p1, getData(e, "unknown")),
  IBCF_C_E = calcPredictionAccuracy(p2, getData(e, "unknown")),
  IBCF_Z_E = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
```

```{r}
c_res <- data.frame(rbind(error_UEUC, error_IEUC))
c_res <- c_res[order(c_res$RMSE ),]
kable(c_res)
```


```{r}
barplot(c_res$RMSE, col = "yellow", main = "Model RMSEs", las = 2, ylab = "RMSE", horiz = FALSE,
        names.arg = rownames(c_res), cex.names=.8)
```


#### Evaluation result
From the above plot, it looks like the performance(prediction accuracy) of User-based Collaborative Fingering  is better than Item-based Collaborative Filtering. 


#### Save pre-trained models
Training a model is time consuming. The model object is stored in .rds format to save time and to be used by the App.


## Technical details

- The models are normalized using normalization settings, like 'center','Z-score'.

- The similarity metric used is Euclidean Distance.

- The prediction we are using on content-based algorithm(System I) is not based on a 'weighted average'.


## Movie Recommendation App

  * It is a shiny app with System I and System II algorithms
  * System I (content based)
    + The app needs to take the input from users of their favorite genres, and display the result
    + System will provide two methods for System I 
  
 * System II (Collaborative Filtering)
   + The app need the users to rate as many movies as possible;
   + System will discover movies for the users by User-based or Item-based Collaborative Filtering models

 * Setting
   + Select the algorithm for System I
   + Select the algorithm for System II


## Running the App

We run the following code on RStudio console using the following command:

**shiny::runApp('app')**


## Conclusion

Multiple models were compared on the dataset, System II Collaborative Filtering is an improvement over the 
System I (content based recommendation system). 



```{r}
Sys.time() - start_time
```

